---
title: "Multi-species functional responses"
format: 
  html:
    self-contained: true
    toc: true
    toc-depth: 3   
    toc-title: "Contents" 
    toc-float: true
    toc-location: left
---

Holling type 2 functional response, see e.g. [Rosenbaum et al. (2024)](https://doi.org/10.1111/2041-210X.14372)

$$F_i(N_1, \dots, N_m)=\frac{a_iN_i}{1+\sum_{k=1}^ma_kh_kN_k}$$

$$
\begin{align}
\frac{dN_1}{dt}=-\frac{a_1N_1}{1+a_1h_1N_1+a_2h_2N_2}P \\
\frac{dN_2}{dt}=-\frac{a_2N_2}{1+a_1h_1N_1+a_2h_2N_2}P
\end{align}
$$

## Dataset

```{r}
#| warning: false
library("BayesFR")
library("ggplot2")
library("cowplot")   # ggplot grids
```

Data from [Colton (1897)](https://doi.org/10.2307/1938361), downloaded from [Novak & Stouffer (2020)](https://doi.org/10.6084/m9.figshare.12830792)


```{r}
df.all = df_Colton_1987_1_ECOLOGY
df.all$Time = df.all$Time/24 # convert units from hours to days
head(df.all)
```

experimental design (combinations of offered prey) full factorial grid, including single-prey experiments ($N_{01}=0$ or $N_{01}=0$) and multi-prey experiments ($N_{01},N_{02}>0$)

```{r}
ggplot(aes(N01,N02), data=df.all) +
  geom_point(size=2) + 
  theme(aspect.ratio=1) 
```

single-prey data species 1 vs species 2: 

```{r}
df.1 = subset(df.all, N02==0)
p1 = ggplot(aes(N01,NE1), data=df.1) +
  geom_jitter(width=0.1, height=0.1, alpha=0.6, size=2) +
  coord_cartesian(xlim=c(0,NA), ylim=c(0,80)) +
  labs(title="Single prey 1")

df.2 = subset(df.all, N01==0)
p2 = ggplot(aes(N02,NE2), data=df.2) +
  geom_jitter(width=0.1, height=0.1, alpha=0.6, size=2) +
  coord_cartesian(xlim=c(0,NA), ylim=c(0,80)) +
  labs(title="Single prey 2")

plot_grid(p1,p2)
```

multi-prey data, proportion of eaten prey $\frac{N_{E1}}{N_{E1}+N_{E2}}$ (y-axis) vs proportion of offered prey $\frac{N_{01}}{N_{01}+N_{02}}$ (x-axis) shows preference for prey species 2. Points below 1-1-line indicate prey species 1 is consumed relatively less often than available in environment 

```{r}
df.all$N01.ratio = df.all$N01/(df.all$N01+df.all$N02)
df.all$NE1.ratio = df.all$NE1/(df.all$NE1+df.all$NE2)

ggplot(aes(N01.ratio,NE1.ratio), data=df.all)+
  geom_point(alpha=0.6, size=2) +
  coord_cartesian(xlim=c(0,1), ylim=c(0,1)) +
  theme(aspect.ratio=1) +
  geom_segment(x=0, y=0, xend=1, yend=1, color="grey40", linetype="dashed") 
```

## Model fitting

Convert wide to long

Remove N0=0

```{r}
df = convert_2sp_to_long(df.all)
str(df)
head(df)
```

```{r}
#| echo: FALSE
#| output: FALSE
FR.formula = bf( NE | trials(N0) ~ MS_Type2H_dyn(N0,N0.alt,ID,P0,Time,a1,a2,h1,h2)/N0,
                 a1+a2+h1+h2 ~ 1, 
                 nl = TRUE)

FR.priors  = c(prior(exponential(1), nlpar="a1", lb=0),
               prior(exponential(1), nlpar="a2", lb=0),
               prior(exponential(10), nlpar="h1", lb=0),
               prior(exponential(10), nlpar="h2", lb=0)
)

fit.1 = brm(FR.formula,
            stanvars = stanvar(scode = MS_Type2H_dyn_code, 
                               block = "functions"),
            family   = binomial(link="identity"),
            prior    = FR.priors,
            data     = df,
            cores    = 4,
            file = "fit_01")
expose_functions(fit.1, vectorize = TRUE)
```

```{r}
#| eval: FALSE
FR.formula = bf( NE|trials(N0) ~ MS_Type2H_dyn(N0,N0.alt,ID,P0,Time,a1,a2,h1,h2)/N0,
                 a1+a2+h1+h2 ~ 1, 
                 nl = TRUE)

FR.priors  = c(prior(exponential(1), nlpar="a1", lb=0),
               prior(exponential(1), nlpar="a2", lb=0),
               prior(exponential(10), nlpar="h1", lb=0),
               prior(exponential(10), nlpar="h2", lb=0)
)

fit.1 = brm(FR.formula,
            stanvars = stanvar(scode = MS_Type2H_dyn_code, 
                               block = "functions"),
            family   = binomial(link="identity"),
            prior    = FR.priors,
            data     = df,
            cores    = 4)
expose_functions(fit.1, vectorize = TRUE)
```


```{r}
summary(fit.1)
```

```{r}
plot(fit.1)
```

## Plotting

```{r}
plot(pp_check(fit.1, ndraws=100))
```

Single prey data

```{r}
#| warning: FALSE
p1 = plot(conditional_effects(fit.1,
                              conditions = data.frame(ID=1, N0.alt=0),
                              int_conditions = data.frame(N0=1:264),
                              effects="N0",
                              ndraws = 500), 
          plot = FALSE)
df.plot = subset(df, ID==1 & N0.alt==0)
p1 = p1[[1]] + geom_point(data = df.plot, 
                          aes(x=N0, y=NE),
                          inherit.aes = FALSE,
                          alpha=0.6, size=2) +
  coord_cartesian(xlim=c(0,NA), ylim=c(0,80)) +  
  labs(title="Single prey 1")

p2 = plot(conditional_effects(fit.1,
                              conditions = data.frame(ID=2, N0.alt=0),
                              int_conditions = data.frame(N0=1:264),
                              effects="N0",
                              ndraws = 500), 
          plot = FALSE)
df.plot = subset(df, ID==2 & N0.alt==0)
p2 = p2[[1]] + geom_point(data = df.plot, 
                          aes(x=N0, y=NE),
                          inherit.aes = FALSE,
                          alpha=0.6, size=2) +
  coord_cartesian(xlim=c(0,NA), ylim=c(0,80)) +  
  labs(title="Single prey 2")

plot_grid(p1,p2)
```

```{r}
#| warning: FALSE
p1 = plot(conditional_effects(fit.1,
                              conditions = data.frame(ID=1),
                              effects="N0:N0.alt",
                              int_conditions = data.frame(N0.alt=c(0,130,260)),
                              ndraws = 500), 
          plot = FALSE)
df.plot = subset(df, ID==1)
p1 = p1[[1]] + geom_point(data = df.plot, 
                          aes(x=N0, y=NE),
                          inherit.aes = FALSE,
                          alpha=0.3, size=2) +
  coord_cartesian(xlim=c(0,NA), ylim=c(0,80)) +
  labs(title="Multi prey 1") +
  theme(legend.position = c(0.2, 0.8))

p2 = plot(conditional_effects(fit.1,
                              conditions = data.frame(ID=2),
                              effects="N0:N0.alt",
                              int_conditions = data.frame(N0.alt=c(0,130,260)),
                              ndraws = 500), 
          plot = FALSE)
df.plot = subset(df, ID==2)
p2 = p2[[1]] + geom_point(data = df.plot, 
                          aes(x=N0, y=NE),
                          inherit.aes = FALSE,
                          alpha=0.3, size=2) +
  coord_cartesian(xlim=c(0,NA), ylim=c(0,80)) +
  labs(title="Multi prey 2") +
  theme(legend.position = c(0.2, 0.8))

plot_grid(p1,p2)
```

```{r}
#| warning: FALSE
# choose level of total prey abundance (constant)
N.total = 250 
df.pred = data.frame(N01 = 1:(N.total-1),
                     N02 = (N.total-1):1)
df.pred$ratio = df.pred$N01/(df.pred$N01+df.pred$N02)

# set up dataframes for posterior predictions
df.pred.1 = data.frame(N0     = df.pred$N01, 
                       N0.alt = df.pred$N02, 
                       ID = 1, P0 = 1, Time = 1)
df.pred.2 = data.frame(N0     = df.pred$N02, 
                       N0.alt = df.pred$N01, 
                       ID = 2, P0 = 1, Time = 1)

# posterior predictions for prey1, prey2, and ratio
post.1 = posterior_epred(fit.1, newdata=df.pred.1)
post.2 = posterior_epred(fit.1, newdata=df.pred.2)
post.ratio = post.1/(post.1+post.2)

# posterior means and credible intervals
df.pred$mean  = colMeans(post.ratio)
df.pred$upper = apply(post.ratio, 2, quantile, probs = 0.975)
df.pred$lower = apply(post.ratio, 2, quantile, probs = 0.025)

ggplot(df.all, aes(x=N01.ratio, y=NE1.ratio)) +
  geom_segment(x=0, y=0, xend=1, yend=1, color="grey30", linetype="dashed") +
  geom_point(alpha=0.4, size=2) +
  geom_ribbon(df.pred, mapping=aes(x=ratio, ymin=lower, ymax=upper), 
              alpha = 0.4, inherit.aes=FALSE) +
  geom_line(df.pred, mapping=aes(x=ratio, y=mean), 
              color = "blue", linewidth = 1, inherit.aes=FALSE) +
  theme(aspect.ratio = 1) 
```

