---
title: "Model checking, model comparison, hypothesis testing: type 2 vs type 3 responses"
format: 
  html:
    self-contained: true
    toc: true
    toc-depth: 3   
    toc-title: "Contents" 
    toc-float: true
    toc-location: left
---

```{r}
#| warning: false
rm(list=ls())
library(BayesFR)
```

Using an example of fitting a type 2 vs. a type 3 FR, this section describes tools for model checking (Does the model describe the data well?), model comparison (Which model describes the data better?), and hypothesis testing (How certain is the model about research statements?). The dataset is from [Sentis et al. (2017)](https://doi.org/10.1111/gcb.13560), was made available in the FoRAGE database ([Uiterwaal et al. 2022](https://doi.org/10.1002/ecy.3706)), and contains three aquatic insect larvae predators feeding on Daphnia prey in two temperature treatments. Eaten prey were not replaced. We are using the predator *S. vulgatum* at 16°C.

```{r}
data(df_Sentis_et_al_2017_GLOBAL_CHANGE_BIOLOGY)
df = subset(df_Sentis_et_al_2017_GLOBAL_CHANGE_BIOLOGY, ID=="Figure 1e")
head(df)
ggplot(aes(N0,NE), data=df) +
  geom_jitter(alpha = 0.5, width=0.5, height=0, size=2.5) +
  coord_cartesian(xlim=c(0,NA), ylim=c(0,NA))
```

## Model checking

We start with fitting a type 2 FR using the dynamical prediction model and a Binomial distribution as described in the last section.
$$\frac{aN}{1+ahN}$$

```{r}
#| eval: false
FR.formula = bf( NE | trials(N0) ~ Type2H_dyn(N0,1.0,1.0,a,h)/N0,
                 a~1, h~1,
                 nl = TRUE)
FR.priors  = c(prior(exponential(1.0), nlpar="a", lb=0),
               prior(exponential(1.0), nlpar="h", lb=0)
)
fit.type2 = brm(FR.formula,
                family   = binomial(link="identity"),
                prior    = FR.priors,
                data     = df,
                cores    = 4,
                stanvars = stanvar(scode=Type2H_dyn_code, block="functions")
)
expose_functions(fit.type2, vectorize=TRUE)
summary(fit.type2)
```

```{r}
#| echo: false
#| output: FALSE
fit.type2 = readRDS(file="fit_type2e.rds")
expose_functions(fit.type2, vectorize=TRUE)
```

```{r}
#| echo: false
#| output: true
summary(fit.type2)
```

The brms package contains some functions for checking how well the model describes the data. We already learned about `conditional_effects()`, which here plots the mean regression curve and its **95% credible intervals**. As default, just the **deterministic model part** is used, which is the mean number of eaten prey `NE` predicted by `Type2H_dyn()`. The uncertainty (credible intervals) comes from the posterior distribution of model parameters `a` and `h` alone. 

```{r}
#| warning: false
plot(conditional_effects(fit.type2), 
     points=TRUE)
```

The statistical model includes not only the deterministic regression, but we assumed that the data is Binomially distributed around these values. This can be specified to plot the **95% prediction intervals**, including also the **stochastic model part**. Note that the curve below looks a bit ragged, since these predictions are integer draws from the Binomial distribution now. The intervals should contain around 95% of the datapoints. In functional response experiments, however, the data is often more noisy than predicted by the Binomial distribution: it is overdispersed. We are not dealing with this right now, but do so later in another chapter.

```{r}
#| warning: false
plot(conditional_effects(fit.type2,
                         method = "posterior_predict"), 
     points=TRUE)
```

An alternative way of displaying model predictions (including both the deterministic and the stochastic model part) against the data are **posterior predictive checks**. `pp_check()` draws a histogram (or rather a density curve) of the observed response values (number of eaten prey) in darkblue: x-axis are response values and y-axis is frequency of occurrence. Every lightblue line depicts the histogram of a replicated dataset from the model, generated with a sample of the parameters' posterior distribution (here, `ndraws=100` replicated datasets). A good alignment of observed and replicated datasets would indicate a good model fit, however there is some mismatch due to data overdispersion. 

```{r}
pp_check(fit.type2, ndraws=100)
```

## Model comparison

We want to know if the data can be described better by an S-shaped type 3 FR. 
$$\frac{bN^2}{1+bhN^2}$$
```{r}
#| eval: false
FR.formula = bf( NE | trials(N0) ~ Type3H_dyn(N0,1.0,1.0,b,h)/N0,
                 b~1, h~1,
                 nl = TRUE)
FR.priors  = c(prior(exponential(1.0), nlpar="b", lb=0),
               prior(exponential(1.0), nlpar="h", lb=0)
)
fit.type3 = brm(FR.formula,
                family   = binomial(link="identity"),
                prior    = FR.priors,
                data     = df,
                cores    = 4,
                stanvars = stanvar(scode=Type3H_dyn_code, block="functions")
)
expose_functions(fit.type3, vectorize=TRUE)
summary(fit.type3)
```

```{r}
#| echo: false
#| output: FALSE
fit.type3 = readRDS(file="fit_type3e.rds")
expose_functions(fit.type3, vectorize=TRUE)
```

```{r}
#| echo: false
#| output: true
summary(fit.type3)
```

Again, we have a look at the diagnostics plots.

```{r}
#| warning: false
plot(conditional_effects(fit.type3), 
     points=TRUE)
plot(conditional_effects(fit.type3,
                         method = "posterior_predict"), 
     points=TRUE)
pp_check(fit.type3, ndraws=100)
```

Like the type 2, the type 3 FR has a bit of a problem with overdispersion. But which one describes the data better? It is possible to compute **R2** values, describing the **amount of explained variation** in the response. However, in this nonlinear modeling framework with non-Gaussian distributions, this measure has limited informative value. E.g., R2 weighs deviations at small `N0` values equally as deviations at large `N0`, and ignores the fact that the Binomial distribution accounts for small variance at lower and larger variance at higher `N0`.

```{r}
bayes_R2(fit.type2)
bayes_R2(fit.type3)
```

Much better suited would be a likelihood-based **information criterion** like AIC. The brms package uses the leave-one-out (**LOO**) cross-validation criterion from the [loo-package](https://cran.r-project.org/web/packages/loo/vignettes/loo2-example.html). Very roughly speaking, LOO accounts for model fit (likelihood) and for model complexity (number of parameters) similarly as AIC. The `LOO()` function produces some infos on both models' LOO values, but since LOO is a relative measure, we are only interested in the difference at the bottom of the output.

```{r}
LOO(fit.type2, fit.type3)
```

The models are ranked by model performance: Here, the type 3 seems to be the better model, with a difference to the type 2 model of `18.1`. However, the difference comes with an associated standard error `18.3` in the second column. Conventionally, one can safely assume that one model outperforms the other one, if the difference is **approximately at least twice as big** as the standard error. Since the estimated difference in model performance is only around one standard error here, the model comparison is inconclusive. 

## Hypothesis testing

While hypotheses can often be tested by comparing one model against the other as described above, we here will calculate posterior probabilities of model parameters using the a generalized type 3 response as an example.

As an alternative to the classical type 2 and type 3 responses, a flexible exponent $q$ is used to make the attack rate $a=bN^q$ density dependent, which defines the functional response
$$\frac{bN^{1+q}}{1+bhN^{1+q}}$$
that generalizes the type 2 ($q=0$) and the type 3 ($q=1$).

We fit a dynamical prediction model `Type3GenH_dyn()` with a free parameter `q`. Unlike the type 2 and the strict type 3 FR, there is no analytical solution for the underlying differential equation, and the prediction model includes a numerical solution to compute number of eaten prey `NE`. We must specify a prior for the additional parameter `q`, where a lower boundary `lb=-1` guarantees that the exponent `1+q` stays positive.


```{r}
#| eval: false
FR.formula = bf( NE | trials(N0) ~ Type3GenH_dyn(N0,1.0,1.0,b,h,q)/N0,
                 b~1, h~1, q~1,
                 nl = TRUE)
FR.priors  = c(prior(exponential(1.0), nlpar="b", lb=0),
               prior(exponential(1.0), nlpar="h", lb=0),
               prior(normal(0,1), nlpar="q", lb=-1)
)
fit.type3gen = brm(FR.formula,
                family   = binomial(link="identity"),
                prior    = FR.priors,
                data     = df,
                cores    = 4,
                stanvars = stanvar(scode=Type3GenH_dyn_code, block="functions")
)
expose_functions(fit.type3gen, vectorize=TRUE)
summary(fit.type3gen)
```

```{r}
#| echo: false
#| output: FALSE
fit.type3gen = readRDS(file="fit_type3gene.rds")
expose_functions(fit.type3gen, vectorize=TRUE)
```

```{r}
#| echo: false
#| output: true
summary(fit.type3gen)
```

```{r}
#| warning: false
plot(conditional_effects(fit.type3gen), 
     points=TRUE)
```

Indeed, the exponent is estimated with a mean of 0.64 and a 95% CI of [0.42, 0.88].

A model comparison reveals that this model performs better than the type 2 model, but is comparable to the type 3 model (difference is small compared to standard error).

```{r}
#| eval: false
LOO(fit.type2, fit.type3, fit.type3gen)
```

```{r}
#| echo: false
#| output: true
loo_compare(loo(fit.type2), loo(fit.type3), loo(fit.type3gen))
```

We examine this a bit more closely by calculating posterior probabilities for the parameter `q_Intercept`. Its 95% credible interval [0.42, 0.88] provides evidence that $0<q<1$, but what if we want to compute posterior probabilities $P(q>0)$ or $P(q<1)$? The brms function `hypothesis()` calculates these, simply by counting how many posterior samples satisfy the condition. It can also be used for quantities derived from the parameters or for predictions, but here we simply ask if the parameter is positive:

```{r}
hypothesis(fit.type3gen, "q_Intercept>0")
p = plot(hypothesis(fit.type3gen, "q_Intercept>0"), plot=FALSE)
p[[1]] + geom_vline(xintercept=0)
```

The information we seek is in the column `Post.Prob`, based on the model fit we are 100% certain that $q>0$. The plot also shows the posterior distribution does not include 0.

Testing against values different from 0, `hypthesis()` re-arranges the output a bit. 

```{r}
hypothesis(fit.type3gen, "q_Intercept<1")
p = plot(hypothesis(fit.type3gen, "q_Intercept<1"), plot=FALSE)
p[[1]] + geom_vline(xintercept=0)
```

Here, we are also certain that $q<1$. 

With the same syntax it is possible to compute, e.g., $P(q>0.1)$ or $P(q<0.9)$, however there is currently no option in `hypothesis()` to compute probabilities of two-sided intervals such as $P(0.1<q<0.9)$. For this, we can extract the posterior samples and calculate it manually.

```{r}
draws = as.data.frame(fit.type3gen)
n = nrow(draws)
head(draws)
```

Each row in the dataframe `draws` contains a multivariate sample of the posterior, and each column contains all posterior samples for a specific model parameter. Notice how the parameter $q$ is named `b_q_Intercept`. brms internally puts `b_` in front of a parameter name if it is an effect size, and `_Intercept` after the parameter name to indicate that it is the effect size of the intercept and not of another predictor (we don't have additional predictors here).

We just have to count how many samples satisfy a condition and divide by the total number of samples, which computes the posterior probability. Multiple conditions can be combined with the logical AND operator `&` to compute, e.g., $P(0.1<q<0.9)=P(q>0.1\ \text{AND}\  q<0.9):$

```{r}
sum(draws$b_q_Intercept>0.1)/n
sum(draws$b_q_Intercept<0.9)/n
sum(draws$b_q_Intercept>0.1 & draws$b_q_Intercept<0.9)/n
```
